Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models

### 2019-06-11 paper.tex
Update on Overleaf.
clean_figs

The general method workflow of our method is illustrated in Figure \ref{fig:method}. In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a spatial generator which depends on its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch). We call these parameters ``generator parameters'' to distinguish them from the standard parameters of the models and to highlight their higher position in the simulation process. The resulting configurations \emph{can be} clustered into qualitative types of spatial patterns. The sensitivity analysis relates the variations in the model's outcomes to how the density spatial distribution was generated and to the patterns of density generated. In particular, we want to emphasize that spatial effects derive not only from grid size or shape effects, but also from the heterogeneity of the distribution of geographical entities (people, housing, networks, etc.) in space. In the models we used as examples, the initial spatial configurations can be either flat or heterogeneous, monocentric or polycentric, based on external databases and on internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
\centering
\includegraphics[width=0.9\textwidth]{figures/SchemaMeta_1.png}
\caption{\textbf{General workflow.}} \label{fig:method}
\end{figure} %
%%%%%%%%%%%%%%%%%%%%%%

@@

where $\mu_{\vec{\alpha}_i}$ are the phase diagrams at given generator parameters $\vec{\alpha}_i$. $d$ is a functional distance that we take simply as Euclidean distance. The internal variabilities are estimated as the variance within each phase diagram $f_{\vec{\alpha_i}}$. For a multi-dimensional phase diagram, we average these relative distances over the components. Given a set of phase diagram to be compared, we will study the distribution of this distance to an arbitrary phase diagram for all diagrams, rather than an aggregated measure which would be similar to global sensitivity methods \citep{saltelli2008global}.

@@

%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
\centering
	\includegraphics[width=\textwidth]{figures/spatialGen.png}
 \caption{\textbf{Four examples of grids produced by the spatial generator.} The lighter the red, the denser the area. Changing the growth rate $N$ allows to have more or less chaotic shapes (two first compared to the two last grids for example) corresponding to different levels of convergence of the model, whereas local radius can be tuned with the interplay of aggregation strength $\alpha$ and diffusion strength $\beta$. Parameter values used here are for the first grid $(\alpha = 0.4,\beta=0.006, N = 25, n_d = 1, t_f = 971)$, the second $(\alpha = 0.4,\beta=0.006, N = 25, n_d = 1, t_f = 176)$, the third $(\alpha = 1.4,\beta=0.045, N = 102, n_d = 2, t_f = 618)$, and the fourth $(\alpha = 1.8,\beta=0.114, N = 108, n_d = 1, t_f = 227)$.}
\label{fig:spatialGen}
\end{figure} %
%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
\centering
	\includegraphics[width=0.9\textwidth]{figures/GridPresentation.png}
 \caption{Correspondence between European urban density structures and grids produced with the spatial generator.} 
\label{fig:densityTypes}
\end{figure} %
%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%
\begin{figure}[!t]
\centering
\includegraphics[width=0.8\textwidth]{figures/relativedistance_metaparams.png}
\caption{\textbf{Relative distances of phase diagrams by initial spatial grids described by their generator parameters.} Relative distance as a function of generator parameters $\alpha$ (strength of preferential attachment) and $\beta$ (strength of diffusion process).}
\label{fig:sugarscape-distance-meta}
\end{figure}
%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%
\begin{figure}[!t]
\centering
\includegraphics[width=0.8\textwidth]{figures/relativedistance_morphspace.png}
\caption{\textbf{Relative distances of phase diagrams to the reference across grids.} Relative distance as a function of two first principal components of the morphological space (see text). Black point correspond to the reference spatial configuration. Green frame and blue frame give respectively the first and second particular phase diagrams shown in Fig.~\ref{fig:sugarscape-phasediagrams}.}
\label{fig:sugarscape-distance-pca}
\end{figure}
%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%
\begin{figure}[!t]
\centering
\includegraphics[width=0.48\textwidth]{figures/phasediagram_id27_maxSugar110}
\includegraphics[width=0.48\textwidth]{figures/phasediagram_id0_maxSugar110}
\caption{\textbf{Examples of phase diagrams.} We show two dimensional phase diagrams on $(P,s_-)$, both at fixed $s_+ = 110$. (Left) Green frame, obtained with $\alpha = 0.79$, $n=2$, $\beta = 0.14$, $N=157$; (Right) Blue frame, obtained with $\alpha = 2.56$, $n=3$, $\beta = 0.13$, $N=128$.}
\label{fig:sugarscape-phasediagrams}
\end{figure}
%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%
\begin{figure}[!t]
	\centering
	\includegraphics[width=0.49\linewidth]{figures/alpha_beta_examples.png}
    \includegraphics[width=0.47\linewidth]{figures/alpha_beta_schemas.png}
    \caption{\textbf{Stylized behavior of the density grid generator.} (Left) Examples of grids with their position in the $(\alpha,\beta)$ parameter space; (Right) Stylized interpretation of the forms obtained as a function of $\alpha$ and $\beta$.}\label{fig:stylized-generator}
\end{figure}
%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%
\begin{figure}[!t]
\centering
\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_metaparams_mean.png}
\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_morphspace_mean.png}
\caption{\textbf{Relative distances of phase diagrams to the reference across grids for the Schelling model.} Each point corresponds to a spatial configuration and colour gives the relative distance to one of the phase diagrams. We show them in the generator parameter space (Left) and in the reduced morphological space (Right).\label{fig:schelling-distance-meta}}
\end{figure}
%%%%%%%%%%%%%

>>>>

The general method workflow of our method is illustrated in Figure \ref{fig:method}. In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a spatial generator which depends on its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch). We call these parameters ``generator parameters'' to distinguish them from the standard parameters of the models and to highlight their higher position in the simulation process. The resulting configurations \emph{can be} clustered into qualitative types of spatial patterns. The sensitivity analysis relates the variations in the model's outcomes to how the density spatial distribution was generated and to the patterns of density generated. In particular, we want to emphasize that spatial effects derive not only from grid size or shape effects, but also from the heterogeneity of the distribution of geographical entities (people, housing, networks, etc.) in space. In the models we used as examples, the initial spatial configurations can be either flat or heterogeneous, monocentric or polycentric, based on external databases and on internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
\centering
%\includegraphics[width=0.9\textwidth]{figuresraw/SchemaMeta_1.png}
\includegraphics[width=0.9\textwidth]{figures/Fig1.png}
\caption{\textbf{General workflow.}} \label{fig:method}
\end{figure} %
%%%%%%%%%%%%%%%%%%%%%%

@@

where $\mu_{\vec{\alpha}_i}$ are the phase diagrams at given generator parameters $\vec{\alpha}_i$. $d$ is a functional distance that we take simply as Euclidean distance. The internal variabilities are estimated as the variance within each phase diagram $\mu_{\vec{\alpha_i}}$. For a multi-dimensional phase diagram, we average these relative distances over the components. Given a set of phase diagram to be compared, we will study the distribution of this distance to an arbitrary phase diagram for all diagrams, rather than an aggregated measure which would be similar to global sensitivity methods \citep{saltelli2008global}.

@@

%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
\centering
	%\includegraphics[width=\textwidth]{figuresraw/spatialGen.png}
	\includegraphics[width=\textwidth]{figures/Fig2.png}
 \caption{\textbf{Four examples of grids produced by the spatial generator.} The lighter the red, the denser the area. Changing the growth rate $N$ allows to have more or less chaotic shapes (two first compared to the two last grids for example) corresponding to different levels of convergence of the model, whereas local radius can be tuned with the interplay of aggregation strength $\alpha$ and diffusion strength $\beta$. Parameter values used here are for the first grid $(\alpha = 0.4,\beta=0.006, N = 25, n_d = 1, t_f = 971)$, the second $(\alpha = 0.4,\beta=0.006, N = 25, n_d = 1, t_f = 176)$, the third $(\alpha = 1.4,\beta=0.045, N = 102, n_d = 2, t_f = 618)$, and the fourth $(\alpha = 1.8,\beta=0.114, N = 108, n_d = 1, t_f = 227)$.}
\label{fig:spatialGen}
\end{figure} %
%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
\centering
	%\includegraphics[width=0.9\textwidth]{figuresraw/GridPresentation.png}
	\includegraphics[width=0.9\textwidth]{figures/Fig3.png}
 \caption{Correspondence between European urban density structures and grids produced with the spatial generator.} 
\label{fig:densityTypes}
\end{figure} %
%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%
\begin{figure}[!t]
\centering
	%\includegraphics[width=0.8\textwidth]{figuresraw/relativedistance_metaparams.png}
	\includegraphics[width=0.8\textwidth]{figures/Fig4.png}
\caption{\textbf{Relative distances of phase diagrams by initial spatial grids described by their generator parameters.} Relative distance as a function of generator parameters $\alpha$ (strength of preferential attachment) and $\beta$ (strength of diffusion process).}
\label{fig:sugarscape-distance-meta}
\end{figure}
%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%
\begin{figure}[!t]
\centering
	%\includegraphics[width=0.8\textwidth]{figuresraw/relativedistance_morphspace.png}
	\includegraphics[width=0.8\textwidth]{figures/Fig5.png}
\caption{\textbf{Relative distances of phase diagrams to the reference across grids.} Relative distance as a function of two first principal components of the morphological space (see text). Black point correspond to the reference spatial configuration. Green frame and blue frame give respectively the first and second particular phase diagrams shown in Fig.~\ref{fig:sugarscape-phasediagrams}.}
\label{fig:sugarscape-distance-pca}
\end{figure}
%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%
\begin{figure}[!t]
\centering
%\includegraphics[width=0.48\textwidth]{figuresraw/phasediagram_id27_maxSugar110}
%\includegraphics[width=0.48\textwidth]{figuresraw/phasediagram_id0_maxSugar110}
	\includegraphics[width=\textwidth]{figures/Fig6.png}
\caption{\textbf{Examples of phase diagrams.} We show two dimensional phase diagrams on $(P,s_-)$, both at fixed $s_+ = 110$. (Left) Green frame, obtained with $\alpha = 0.79$, $n=2$, $\beta = 0.14$, $N=157$; (Right) Blue frame, obtained with $\alpha = 2.56$, $n=3$, $\beta = 0.13$, $N=128$.}
\label{fig:sugarscape-phasediagrams}
\end{figure}
%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%
\begin{figure}[!t]
	\centering
	%\includegraphics[width=0.49\linewidth]{figuresraw/alpha_beta_examples.png}
    %\includegraphics[width=0.47\linewidth]{figuresraw/alpha_beta_schemas.png}
    \includegraphics[width=\textwidth]{figures/Fig7.png}
    \caption{\textbf{Stylized behavior of the density grid generator.} (Left) Examples of grids with their position in the $(\alpha,\beta)$ parameter space; (Right) Stylized interpretation of the forms obtained as a function of $\alpha$ and $\beta$.}\label{fig:stylized-generator}
\end{figure}
%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%
\begin{figure}[!t]
\centering
%\includegraphics[width=0.48\textwidth]{figuresraw/schelling-relativedistance_metaparams_mean.png}
%\includegraphics[width=0.48\textwidth]{figuresraw/schelling-relativedistance_morphspace_mean.png}
\includegraphics[width=\textwidth]{figures/Fig8.png}
\caption{\textbf{Relative distances of phase diagrams to the reference across grids for the Schelling model.} Each point corresponds to a spatial configuration and colour gives the relative distance to one of the phase diagrams. We show them in the generator parameter space (Left) and in the reduced morphological space (Right).\label{fig:schelling-distance-meta}}
\end{figure}
%%%%%%%%%%%%%
